Liveness-Based Garbage Collection for Lazy Languages

### 2015-03-13 paper/paper.tex
update paper.tex
Updated section on liveness analysis for lazy languages

\documentclass{sig-alternate}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage{algorithmic}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it 
      %probably doesn't scale anyway.
%\newcommand{\comment}[1]{}
\definecolor{Myblue}{rgb}{0.0,0.0,0.5}
\newcommand{\comment}[1]{{\color{Myblue}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\scmin} {\mbox{\sf\em in}}
\newcommand{\scmnull}{\mbox{\sf\em null?}}
\newcommand{\scmpair}{\mbox{\sf\em pair?}}
\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
%\nonstopmode
\def\myvec{\mathaccent"017E } % wretched springer mode redefines \vec as bold!
\newcommand{\stk}{\mbox{S}}       % stack
\newcommand{\ID}{\mbox{$\mathbf{ id}$}} % identity function
\newcommand{\bang}{\mbox{\sc bang}}

\begin{document}
%
% --- Author Metadata here ---
\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

@@

Consider a call to $\pf$  with a demand $\sigma$. For obvious reasons,
this is also the demand on the return values $u$ and $y$, and thus the
liveness  at  the  program  points  $\pi_7$ and  $\pi_8$  re  $\lbrace
\pu.\sigma \rbrace$ and $\{\py.\sigma\}$.  The liveness at $\pi_7$ and
$\pi_8$ are propagated backwards as in traditional liveness analysis.

@@

  $\pi_6$, the point of its last use. In a lazy language, however, the

@@

  To address the first issue, we  regard the body of the function as a
  tree and consider paths from the root of this tree to program points
  which     trigger     a     evaluation---\SIF\    conditions     and
  \RETURN\  expressions. We call  such paths  {\em E-paths}.   For the
  program  shown, there  are three  such paths  $EP_1 =  \pi_1, \pi_2,
  \pi_3,  \pi_4, \pi_5$, $EP_1  = \pi_1,  \pi_2, \pi_3,  \pi_4, \pi_6,
  \pi_7, \pi_8$, and $EP_2 =  \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.  The
  analysis consists of two maps:  The map $\cal{M}$ maps a return path
  to a liveness  environment that is common to  all the program points
  in the path. And a second  map $\cal{P}$ maps a program point to the
  set of  return paths on  which the point  lies.  If a  program point
  happens  to fall  on  more than  one  such path,  then the  liveness
  environment of the  program point is the variable-wise  union of the
  liveness  environments  of the  individual  paths.   In the  example
  program,  ${\cal  M}(EP_1)=  \ldots$,  ${\cal  M}(EP_2)=\ldots$  and
  ${\cal M}(EP_3)= \ldots$.  Since  ${\cal P}(pi_4)$ is $\lbrace EP_1,
  EP_2, EP_3 \rbrace$, the liveness of $pi_4$ is \ldots.

>>>>

\documentclass{sig-alternate}
\usepackage{etex}
\usepackage{float}
\usepackage{mathrsfs}
\usepackage{booktabs}
\usepackage{boxedminipage}
\usepackage[T1]{fontenc}
\usepackage[pdf]{pstricks}
\usepackage{epsfig}
\usepackage{multirow}
\usepackage{url}
\usepackage[normalem]{ulem}
\usepackage{schemepgm}
\usepackage{graphicx}
\usepackage{graphpap}
\usepackage{tabularx}
\usepackage{algorithmic}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{pstricks}
\usepackage{pst-text}
\usepackage{pst-node}
\usepackage{pst-tree}
\usepackage{pst-rel-points}
\usepackage{bcprules}
\input{fh_defs}
\def\drawplusplus#1#2#3{\hbox to 0pt{\hbox to #1{\hfill\vrule height #3 depth
      0pt width #2\hfill\vrule height #3 depth 0pt width #2\hfill
      }}\vbox to #3{\vfill\hrule height #2 depth 0pt width
      #1 \vfill}}
      %Poor man's typography
\def\concat{\mathrel{\drawplusplus {12pt}{0.4pt}{5pt}}}
      %It would be better to specify these in font-relative measures, but it 
      %probably doesn't scale anyway.
%\newcommand{\comment}[1]{}
\definecolor{Myblue}{rgb}{0.0,0.0,0.5}
\newcommand{\comment}[1]{{\color{Myblue}{\medskip \hrule\medskip
    #1 \medskip \hrule \medskip}}}
\newcommand{\scmin} {\mbox{\sf\em in}}
\newcommand{\scmnull}{\mbox{\sf\em null?}}
\newcommand{\scmpair}{\mbox{\sf\em pair?}}
\newcommand{\scmprim}{\ensuremath{\mathsf{+}}}
\newcommand{\sh}[1]{{\colorbox{gray!20}{\framebox{$#1$}}}}
\def\myvec{\mathaccent"017E } % wretched springer mode redefines \vec as bold!
\newcommand{\stk}{\mbox{S}}       % stack
\newcommand{\ID}{\mbox{$\mathbf{ id}$}} % identity function
\newcommand{\bang}{\mbox{\sc bang}}

\begin{document}
% --- Author Metadata here ---
\conferenceinfo{WOODSTOCK}{'97 El Paso, Texas USA}
%\CopyrightYear{2007} % Allows default copyright year (20XX) to be over-ridden - IF NEED BE.
%\crdata{0-12345-67-8/90/01}  % Allows default copyright data (0-89791-88-6/97/05) to be over-ridden - IF NEED BE.
% --- End of Author Metadata ---

@@

The concept of demand and liveness for lazy languages are almost similar to their 
counterparts in the eager language. The important distinction comes from the fact 
that in eager languages, variables can point only to data values (either atomic values or cons cells). 
Hence any demand on the variable can easily be transferred onto the data being pointed. 
In lazy languages, a variable can either point to a fully evaluated data value or it 
could point to a closure which on evaluation generates the required data value. For example, 
consider the statement {\tt let \py~ $\leftarrow $ (\CDR~\px)}. In case of an eager language 
the variable \py~ points to  the \CDR~ part of \px~ which is guaranteed to be a cons cell. 
Hence any demand on \py~ can be immediately transferred to the \CDR~ part of \px. In case of 
a lazy language, \px~ might not be evaluated and  hence pointing to a closure. In our 
example, it points to the closure corresponding to (\CONS~1~\px). It does not make any sense 
to transfer the demand onto the \CDR~ part of \px~ as it still does not exist. In such a case 
we treat the closure as a demand transformer and tranform the demand on \py~ to the arguments 
of the closure pointed by \px. This procedure is similar to the process of transferring the demand 
from the return value of a function to the arguments of the function call in an eager language. In case 
of lazy languages this process cannot be done statically as we cannot statically determine whether a variable points to 
a data value or a closure. Our method approximates the liveness of closure arguments at runtime 
by tagging extra information with each closure. Since the arguments of the closures could themselves be closures in turn, 
we keep recursively applying the same procedure until we hit a data value. The only case where we do not 
follow this procedure is if the demand on a particular closure is null. In such a case we safely collect 
the entire closure as garbage.

We now describe the process of finding out the liveness 
of each variable defined in the function $\pf$ assuming a symbolic demand $\sigma$ on it. 
For obvious reasons, this is also the demand on the return values $u$ and $y$, and thus the
liveness  at  the  program  points  $\pi_7$ and  $\pi_8$  are  $\lbrace
\pu.\sigma \rbrace$ and $\{\py.\sigma\}$.  The liveness at $\pi_7$ and
$\pi_8$ are propagated backwards as in traditional liveness analysis.

@@

  $\pi_7$, the point of its last use. In a lazy language, however, the

@@

  To address the first issue, we  regard the body of the function as a
  tree and consider paths from the root of this tree to program points
  which     trigger     a     evaluation---\SIF\    conditions     and
  \RETURN\  expressions. We call  such paths  {\em E-paths}.   For the
  program  shown, there  are three  such paths  $EP_1 =  \pi_1, \pi_2,
  \pi_3,  \pi_4, \pi_5$, $EP_2  = \pi_1,  \pi_2, \pi_3,  \pi_4, \pi_6,
  \pi_7, \pi_8$, and $EP_3 =  \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.  The
  analysis consists of two maps:  The map $\cal{M}$ maps a return path
  to a liveness  environment that is common to  all the program points
  in the path. And a second  map $\cal{P}$ maps a program point to the
  set of  return paths on  which the point  lies.  If a  program point
  happens  to fall  on  more than  one  such path,  then the  liveness
  environment of the  program point is the variable-wise  union of the
  liveness  environments  of the  individual  paths.   In the  example
  program,  ${\cal  M}(EP_1)=  \ldots$,  ${\cal  M}(EP_2)=\ldots$  and
  ${\cal M}(EP_3)= \ldots$.  Since  ${\cal P}(pi_4)$ is $\lbrace EP_1,
  EP_2, EP_3 \rbrace$, the liveness of $pi_4$ is \ldots.
